The data for this assignment was downloaded from the <link provided in the assignment>[link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) into the working directory for processing with the following R code.

The data were read into the working directory for processing.  
```{r echo = TRUE}
ActMon <- read.csv("activity.csv")
```  

The total number of steps taken per day were calculated and plotted as a histogram with 10 bins.
```{r echo = TRUE}
steps.date <- tapply(ActMon$steps, ActMon$date, FUN = "sum")
hist(steps.date, breaks = 10,
     xlab = "steps per day",
     main = "Frequency of Daily Step Count")
```

The mean and median steps per day were calculated as: mean = `r format(mean(steps.date, na.rm = TRUE), scientific = FALSE)`, median =`r median(steps.date, na.rm = TRUE)`.

Averaging steps per 5-minute interval across all days was plotted to view average activity across the day.
```{r echo = TRUE}
steps.int <- tapply(ActMon$steps, ActMon$interval, FUN = "mean", na.rm = TRUE)
ints <- unique(ActMon$interval)
plot(ints, steps.int, type = "l",
     xlab = "daily 5-minute interval",
     ylab = "steps taken",
     main = "Average Steps Taken Every 5 Minutes")
```

The interval with the highest number of steps was identified as `r which.max(steps.int)`, which translates to mean interval #104 that occurred between 8:35 - 8:40am.

The number of missing values in the data was calculated as `r sum(is.na(ActMon$steps))`, which represented `r round(sum(is.na(ActMon$steps))/length(ActMon$steps)*100, digits = 1)` percent of the total data.

Missing values were imputed using the mean value for the interval across all days without missing values for that interval and added into a replicate data set.
```{r echo = TRUE}
library(dplyr)
ActMon2 <- ActMon
ActMon2 <- ActMon2 %>%
    group_by(interval) %>%
    mutate(steps=replace(steps, is.na(steps), mean(steps, na.rm=TRUE)))
```

Histograms showing the frequency of steps per day were regenerated with the imputed data.
```{r echo = TRUE}
steps2.date <- tapply(ActMon2$steps, ActMon2$date, FUN = "sum")
hist(steps2.date, breaks = 10,
     xlab = "steps per day",
     main = "Frequency of Daily Step Count")
```

The mean and median were recalculated after imputing the missing data as: mean = `r format(mean(steps2.date), scientific = FALSE)`, median = `r format(median(steps2.date), scientific = FALSE)`

As expected, there was no change in the mean, since the mean of the intervals was used to impute the new values, and there was a small change in the median, since a fractional value was imputed into the data set.

For further analysis, days were identified as weekdays or weekend days with an additional factor column added to the data set
```{r echo = TRUE}
## convert date column to POSITxt
library(lubridate)
library(plyr)
ActMon2$date <- ymd(ActMon2$date)
## add day factor
ActMon2$day.type <- as.factor(weekdays(ActMon2$date))
## convert days to weekday or weekend factors
ActMon2$day.type <- revalue(ActMon2$day.type, 
        c("Monday"="Weekday", "Tuesday"="Weekday", "Wednesday"="Weekday", 
        "Thursday"="Weekday", "Friday"="Weekday","Saturday"="Weekend","Sunday"="Weekend"))
```

The average steps per day were recalculated separately for weekdays and weekend days.
```{r echo = TRUE}
ActMon.wkday <- subset(ActMon2, day.type=="Weekday")
ActMon.wkend <- subset(ActMon2, day.type=="Weekend")
## calculate average steps by interval for weekday and weekend separately
steps.wkday <- tapply(ActMon.wkday$steps, ActMon.wkday$interval, FUN = "mean")
steps.wkend <- tapply(ActMon.wkend$steps, ActMon.wkend$interval, FUN = "mean")
```

The step data by day type was replotted to view average daily activity across the 5-minute intervals separately by weekday and weekend day.
```{r echo = TRUE}
par(mfrow=c(2,1))
plot(ints, steps.wkday, type = "l",
     xlab = "daily 5-minute interval",
     ylab = "steps taken",
     main = "Average Weekday Steps Taken Every 5 Minutes")
plot(ints, steps.wkend, type = "l",
     xlab = "daily 5-minute interval",
     ylab = "steps taken",
     main = "Average Weekend Steps Taken Every 5 Minutes")
```

These graphs show greater activity during mid-day time periods on the weekend then during weekdays for this individual.
